---
title: "DATA471 Assignment 4"
author: 'Loren Whithair (Student ID: 300 434 521)'
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1
## Regarding the computation of order statistics

Hyndman and Fan (1998) proposed 9 ways to compute the quantile over a set of sample data.

Generally speaking, to obtain the quantile p for ordered data x of size N, the function will look at `x[floor(p*N)]` and `x[floor(p*N) + 1]`, and use one of two general methods to compute the quantile:

1. a discrete method that picks a value between these two ordered values in x set by another constant
1. a continuous interpolation method that "smooths" the quantile values when `x[floor(p*N)] <> x[floor(p*N) + 1]`

Extreme values of p(very close to 0 or 1) are handled explicitly to be x[1] and x[N] respectively.

Wicklin shows that for n >= 100, the difference between the methods are minimal. Differences are noticeable for smaller datasets however, and so in these scenarios it's important to report which one you used.

My version of R uses type=7, which is one of the continuous interpolation methods.

I would use the default unless there was strong reasoning for otherwise (e.g. small dataset where a discrete method is more appropriate given the context). Using the default is common practice and therefore makes it quicker simpler for other statisticians to understand outputs.


## First-order bootstrap bias corrected estimator

First, we create the function that allows us to compute our original estimator.

```{r e1_lambda_iqr}
lambda_iqr <- function(data){
  return (log(3) / IQR(data))
}
set.seed(12345)
# an example:
x <- rexp(11, rate=1)
print(lambda_iqr(x))
```

We then create a function that uses bootstrapping to estimated the expected value.
```{r e1_estimated_value}

bootstrap_lambda <- function(data){
  boot_output <- rep(0, 200)
  for(l in 1:200){
    boot_output[l] <- lambda_iqr(sample(data, size=length(data), replace=TRUE))
  }
  return(boot_output)
}
print(mean(bootstrap_lambda(x)))
```

Lastly, create the function that returns the bootstrap bias corrected estimator for $\lambda_{IQR}$:

```{r e1_lambda_tilda}
lambda_tilda <- function(data){
  (2*lambda_iqr(data)) - mean(bootstrap_lambda(x))
}
```

Checking it works:

```{r e1_test}
set.seed(1234567)
x <- rexp(100, rate=1)
paste0("Lambda IQR: ",lambda_iqr(x))
paste0("Lambda Tilda: ", lambda_tilda(x))
```
Note that the inter-quartile range for a bootstrapped sample can be zero. This is more likely to occur for smaller x and larger B (bootstrapping iterations). This can make our parameter estimator infinite (divide by zero issue).

## Monte Carlo Experiment

Firstly, we need to instantiate the final estimator function, based on the maximum likelihood:

```{r lambda_ml}
lambda_ml <- function(data){
  return(mean(data))
}
# checking it works:
lambda_ml(rexp(10, rate=1))
```
Next, we perform the Monte Carlo experiment:

```{r monte_carlo}

reps <- 5000 # the number of replications per lambda, n combo

results <- data.frame()

for (lam in c(.1,1,10)){ #all lambda (true rate of the exponential) we are testing
  for (n in c(5,30,100)){# all sample sizes we are testing
    theta.ml <- rep(NA, reps)
    theta.iqr <- rep(NA, reps)
    theta.iqr_tilda <- rep(NA, reps)
    for (r in 1:reps){
      x <- rexp(n, rate=lam)
      theta.ml[r] <- lambda_ml(x)
      theta.iqr[r] <- lambda_iqr(x)
      theta.iqr_tilda[r] <- lambda_tilda(x) # lambda_iqr adjusted for bias with bootstrapping
    }

    # calculate the bias of each theta, B
    # - uses law of large numbers (LLM) to approximate expected value as mean of multiple samples
    # - lam is the true statistics that we are trying to estimate
    bias.ml <- mean(theta.ml) - lam 
    bias.iqr <- mean(theta.iqr) - lam 
    bias.iqr_tilda <- mean(theta.iqr_tilda) - lam 
    
    # calculate the variance of each theta
    var.ml <- var(theta.ml)
    var.iqr <- var(theta.iqr)
    var.iqr_tilda <- var(theta.iqr_tilda)

    # calculate the MSE
    mse.ml <- bias.ml^2 + var.ml
    mse.iqr <- bias.iqr^2 + var.iqr
    mse.iqr_tilda <- bias.iqr_tilda^2 + var.iqr_tilda
    
    # store the values
    results <- rbind(results, data.frame(
      lambda = lam,
      n = n,
      MSE_ml = mse.ml,
      MSE_iqr = mse.iqr,
      MSE_iqr_tilda = mse.iqr_tilda
    ))
  }
}
```

#### Results of the Monte Caro 
