---
title: "DATA471 Assignment 4"
author: 'Loren Whithair (Student ID: 300 434 521)'
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1
## Regarding the computation of order statistics

Hyndman and Fan (1998) proposed 9 ways to compute the quantile over a set of sample data.

Generally speaking, to obtain the quantile p for ordered data x of size N, the function will look at `x[floor(p*N)]` and `x[floor(p*N) + 1]`, and use one of two general methods to compute the quantile:

1. a discrete method that picks a value between these two ordered values in x set by another constant
1. a continuous interpolation method that "smooths" the quantile values when `x[floor(p*N)] <> x[floor(p*N) + 1]`

Extreme values of p(very close to 0 or 1) are handled explicitly to be x[1] and x[N] respectively.

Wicklin shows that for n >= 100, the difference between the methods are minimal. Differences are noticeable for smaller datasets however, and so in these scenarios it's important to report which one you used.

My version of R uses type=7, which is one of the continuous interpolation methods.

I would use the default unless there was strong reasoning for otherwise (e.g. small dataset where a discrete method is more appropriate given the context). Using the default is common practice and therefore makes it quicker simpler for other statisticians to understand outputs.


## First-order bootstrap bias corrected estimator

First, we create the function that allows us to compute our original estimator.


```{r e1_lambda_iqr}
lambda_iqr <- function(X){
  q <- quantile(X)
  return (log(3) / (q["75%"] - q["25%"]))
}